---
title: 'Lab assignment 3'
author: ""
output:
  html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

## Work breakdown:

#### Next, generate the messages

# Part I: Parameter estimation

### Task 1

### Task 2

The Poisson distribution $P(\theta)$ has mean $E[X] = \theta$ and
variance $Var[X] = \theta$. Our team id number is 23.\
We aim to evaluate different methods of constructing confidence
intervals for the parameter θ of a Poisson distribution. The key
questions are:\
1)How well do different CI methods capture the true parameter 𝜃 with the
prescribed confidence level 1-a?\
2)How do the lengths of the intervals compare in terms of precision?\
3)Which method is the most practical and reliable for
Poisson-distributed data?\
Our main question is how well do the three proposed confidence interval
constructions perform in capturing the true parameter $\theta$ for
samples drawn from a Poisson distribution?\
We will assess this by estimating the coverage probability (the fraction
of times the interval contains $\theta$) and the precision (the average
length of the interval) for different sample sizes $n$.\

1.Known Variance: Assuming $\theta$ is known for the variance
$\sigma^2 = \theta$.The $1-\alpha$ confidence interval for $\theta$ is
derived from
$|Z| = |\frac{\sqrt{n}(\bar{X}_n - \theta)}{\sqrt{\theta}}| \le z_{\alpha/2}$,
where $z_{\alpha/2}$ is the $(1-\alpha/2)$-th quantile of
$N(0, 1)$.$$\left[\bar{X}_n - z_{\alpha/2}\frac{\sqrt{\theta}}{\sqrt{n}}, \quad \bar{X}_n + z_{\alpha/2}\frac{\sqrt{\theta}}{\sqrt{n}}\right]$$\
2.Parameter-Independent: From Part (2), the defining inequality is
$\frac{n(\bar{X}_n - \theta)^2}{\theta} \le z_{\alpha/2}^2$. This is a
quadratic inequality in $\sqrt{\theta}$. Solving
$\frac{n(\bar{X}_n - \theta)^2}{\theta} = z_{\alpha/2}^2$ for $\theta$
yields the
limits:$$\theta_{\pm} = \bar{X}_n + \frac{z_{\alpha/2}^2}{2n} \pm \frac{z_{\alpha/2}}{\sqrt{n}} \sqrt{\bar{X}_n + \frac{z_{\alpha/2}^2}{4n}}$$The
$1-\alpha$ confidence interval
is:$$\left[\bar{X}_n + \frac{z_{\alpha/2}^2}{2n} - \frac{z_{\alpha/2}}{\sqrt{n}} \sqrt{\bar{X}_n + \frac{z_{\alpha/2}^2}{4n}}, \quad \bar{X}_n + \frac{z_{\alpha/2}^2}{2n} + \frac{z_{\alpha/2}}{\sqrt{n}} \sqrt{\bar{X}_n + \frac{z_{\alpha/2}^2}{4n}}\right]$$
3.

```{r}
id <- 23
set.seed(id)
theta <- id/10
M <- 10000
N_small <- 50
N_large <- 200
alpha_vals <- c(0.1, 0.05, 0.01)

run_simulation <- function(n, theta, M, alpha_vals) {
    x <- matrix(rpois(n * M, lambda = theta), nrow = n)

    sample_mean <- colMeans(x)

    results <- list()
    for (alpha in alpha_vals) {
        z_alpha2 <- qnorm(1 - alpha / 2)
        
        # --- Part (2): Theoretical CI (known variance theta, for comparison) ---
        # CI_2: [X_bar - z_a/2 * sqrt(theta/n), X_bar + z_a/2 * sqrt(theta/n)]
        L2 <- sample_mean - z_alpha2 * sqrt(theta / n)
        U2 <- sample_mean + z_alpha2 * sqrt(theta / n)
        coverage2 <- mean(theta >= L2 & theta <= U2)
        length2 <- 2 * z_alpha2 * sqrt(theta / n) # CI length is constant here
        
        # --- Part (3): Parameter-Independent CI (Solving quadratic) ---
        # CI_3: [X_bar + z^2/(2n) +/- (z/sqrt(n)) * sqrt(X_bar + z^2/(4n))]
        a <- z_alpha2^2 / (2 * n)
        b <- (z_alpha2 / sqrt(n))
        
        L3 <- sample_mean + a - b * sqrt(sample_mean + a / 2) # Note: a/2 = z^2/(4n)
        U3 <- sample_mean + a + b * sqrt(sample_mean + a / 2)
        
        coverage3 <- mean(theta >= L3 & theta <= U3)
        length3 <- U3 - L3 # CI length is variable
        
        # --- Part (4): Estimated Variance (Wald/Plug-in CI) ---
        # CI_4: [X_bar - z_a/2 * sqrt(X_bar/n), X_bar + z_a/2 * sqrt(X_bar/n)]
        # Note: We cap X_bar at a tiny positive value to avoid sqrt(0) error if all samples are 0.
        sample_sd <- apply(x, 2, sd)       # sample SD for each simulation
        se <- sample_sd / sqrt(n)          # standard error of the mean
        t_alpha2 <- qt(1 - alpha/2, df = n-1)  # t quantile

        L4 <- sample_mean - t_alpha2 * se
        U4 <- sample_mean + t_alpha2 * se

        coverage4 <- mean(theta >= L4 & theta <= U4)
        length4 <- U4 - L4
        
        results[[paste0("alpha_", alpha)]] <- list(
            Coverage = c(CI2=coverage2, CI3=coverage3, CI4=coverage4),
            Avg_Length = c(CI2=mean(length2), CI3=mean(length3), CI4=mean(length4))
        )
    }
    return(results)
}

results_N50 <- run_simulation(N_small, theta, M, alpha_vals)
results_N200 <- run_simulation(N_large, theta, M, alpha_vals)

# --- Output the results ---
cat("\n## 📊 Simulation Results (theta = 2.3) ##\n")
cat("\n### 🔬 Sample Size n = 50: Coverage Probability ###\n")
print(do.call(rbind, lapply(results_N50, function(x) x$Coverage)))

cat("\n### 🔬 Sample Size n = 50: Average CI Length ###\n")
print(do.call(rbind, lapply(results_N50, function(x) x$Avg_Length)))

cat("\n### 🔬 Sample Size n = 200: Coverage Probability ###\n")
print(do.call(rbind, lapply(results_N200, function(x) x$Coverage)))

cat("\n### 🔬 Sample Size n = 200: Average CI Length ###\n")
print(do.call(rbind, lapply(results_N200, function(x) x$Avg_Length)))

# --- Plotting the sample means (Illustrative Statistics) ---
cat("\n## 📈 Illustrative Statistics and Histograms ##\n")
cat(paste("True Parameter theta:", theta, "\n"))
# Note: Re-running the simulation setup to get fresh data for descriptive stats
x_n50_all <- matrix(rpois(N_small * M, lambda = theta), nrow = N_small)
x_n200_all <- matrix(rpois(N_large * M, lambda = theta), nrow = N_large)
cat(paste("Mean of Sample Means (n=50):", mean(colMeans(x_n50_all)), "\n"))
cat(paste("Mean of Sample Means (n=200):", mean(colMeans(x_n200_all)), "\n"))

# Generate sample means for plotting
x_n50 <- colMeans(x_n50_all)
x_n200 <- colMeans(x_n200_all)
```

### Justification

The foundation of all three confidence intervals is the Central Limit
Theorem
(CLT).$$\frac{\sqrt{n}(\bar{X}_n - \theta)}{\sigma} \stackrel{d}{\to} N(0, 1) \quad \text{as } n \to \infty$$For
the Poisson distribution $P(\theta)$, the standard deviation is
$\sigma = \sqrt{\theta}$.Therefore, the statistic
$Z = \frac{\sqrt{n}(\bar{X}_n - \theta)}{\sqrt{\theta}}$ is
approximately standard normal $N(0, 1)$ for large $n$.\
1)2nd method directly uses the statistic $Z$ and the approximation
$P(|Z| \le z_{\alpha/2}) \approx 1-\alpha$. The coverage is expected to
be close to $1-\alpha$ due to the CLT. The simulation confirms this,
with coverages very close to the target.\
2)4rd method liminates θ from the CI by solving a quadratic inequality;
this ensures coverage without knowing the true variance.\
3)4th method Uses the sample standard deviation and the t-distribution,
making it practical for real-world scenarios where θ is unknown.

### Conclusion

# Part II: Unbiasedness of Estimators

### Task 3

```{r}
set.seed(42)
n <- 100
mu <- 10
sigma_squared <- 4
sigma <- sqrt(sigma_squared)
dataset <- rnorm(n, mean = mu, sd = sigma)
head(dataset)
## [1] 12.741917 8.870604 10.726257 11.265725 10.808537 9.787751
cat("Population Mean (mu):", mu, "\n")
## Population Mean (mu): 10
cat("Population Variance (sigma_squared):", sigma_squared, "\n")
## Population Variance (sigma_squared): 4
sample_mean <- mean(dataset)
sample_variance <- var(dataset)
cat("Sample Mean:", sample_mean, "\n")
## Sample Mean: 10.06503
cat("Sample Variance:", sample_variance, "\n")
## Sample Variance: 4.337697
```

## Task a)

```{r}

sigma_n_sq_single <- (1/n) * sum((dataset - sample_mean)^2)
sigma_n_minus_1_sq_single <- (1/(n-1)) * sum((dataset - sample_mean)^2)

cat("Task a) \n")
cat(paste("True Population Variance (sigma^2):", sigma_squared, "\n"))
cat(paste("Biased Estimator (sigma_n^2):", sigma_n_sq_single, "\n"))
cat(paste("Unbiased Estimator (sigma_n-1^2):", sigma_n_minus_1_sq_single, "\n"))
```

## Task b)

```{r}


N_vals <- c(10, 50, 100, 1000)

results_b <- data.frame(
    n = integer(),
    sigma_n_sq = numeric(),
    sigma_n_minus_1_sq = numeric()
)

for (current_n in N_vals) {
    current_dataset <- rnorm(current_n, mean = mu, sd = sigma)

    sum_sq_dev <- sum((current_dataset - current_sample_mean)^2)

    # Calculate Biased Estimator (sigma_n^2)
    sigma_n_sq_calc <- sum_sq_dev / current_n

    # Calculate Unbiased Estimator (sigma_n-1^2)
    sigma_n_minus_1_sq_calc <- sum_sq_dev / (current_n - 1)

    results_b <- rbind(results_b, data.frame(
        n = current_n,
        sigma_n_sq = sigma_n_sq_calc,
        sigma_n_minus_1_sq = sigma_n_minus_1_sq_calc
    ))
}
print(results_b)

# In part b) sigma_n_sq is closer to theoretical sigma_sq when the sample variance > theoretical one.
# That`s because sigma_n_sq is always less than sigma_n_minus_1_sq
```

## Task c)

```{r}


run_bias_simulation <- function(n, mu, sigma_squared, M) {
    samples_matrix <- matrix(rnorm(n * M, mean = mu, sd = sqrt(sigma_squared)), nrow = n)
    sample_means <- colMeans(samples_matrix)

    # Calculate Sum of Squared Deviations (SSD)
    deviations <- samples_matrix - matrix(sample_means, nrow = n, ncol = M, byrow = TRUE)
    sum_sq_dev <- colSums(deviations^2)

    # Biased Estimator (divides by n)
    sigma_n_sq_vals <- sum_sq_dev / n
    # Unbiased Estimator (divides by n-1)
    sigma_n_minus_1_sq_vals <- sum_sq_dev / (n - 1)

    # Calculate Expected Values (Average over M simulations)
    E_sigma_n_sq <- mean(sigma_n_sq_vals)
    E_sigma_n_minus_1_sq <- mean(sigma_n_minus_1_sq_vals)

    # Bias = Expected Value - True Parameter
    Bias_n <- E_sigma_n_sq - sigma_squared
    Bias_n_minus_1 <- E_sigma_n_minus_1_sq - sigma_squared
    return(data.frame(
        n = n,
        True_Var = sigma_squared,
        E_sigma_n_sq = E_sigma_n_sq,
        Bias_sigma_n_sq = Bias_n,
        E_sigma_n_minus_1_sq = E_sigma_n_minus_1_sq,
        Bias_sigma_n_minus_1_sq = Bias_n_minus_1
    ))
}

M_sim <- 10000

results_list <- lapply(N_vals, run_bias_simulation, mu=mu, sigma_squared=sigma_squared, M=M_sim)
final_results_c <- do.call(rbind, results_list)

print(final_results_c)

```

## Task d)

We observe different behaviours of estimators:

The Unbiased Estimator (sigma_n-1\^2): The expected value
E[sigma_n-1\^2] remains consistently close to the true population
variance (sigma\^2 = 4) for all sample sizes (n=10, 50, 100, 1000). The
calculated bias is small regardless of n. This confirms that
sigma_n-1\^2 is an unbiased estimator.

The Biased Estimator (sigma_n\^2): For small sample sizes (e.g., n=10),
this estimator significantly underestimates the true variance
(E[sigma_10\^2] is approx 3.6, which is 10% lower than 4). As n
increases, the underestimation becomes smaller. At n=10, the bias is
large (approx -0.4), but at n=1000, the bias is very small (approx
-0.004), making the estimate nearly identical to the unbiased one.

Mathematical Reason: This happens because the ratio (n-1)/n approaches 1
as n approaches infinity. Therefore, the limit of E[sigma_n\^2] as n
approaches infinity equals sigma\^2.

Conclusion: While sigma_n\^2 is biased for finite samples, it is
asymptotically unbiased. The difference between the two estimators is
critical for small datasets but becomes small for very large datasets.
For statistics sigma_n-1\^2 is always preferred.

## Task e)

We aim to derive the expected values for both estimators:
$$\sigma_n^2 = \frac{1}{n} \sum_{i=1}^n (X_i - \overline{X})^2 \quad \text{and} \quad \sigma_{n-1}^2 = \frac{1}{n-1} \sum_{i=1}^n (X_i - \overline{X})^2$$

First, let us analyze the expected value of the sum of squared
deviations: $E\left[ \sum_{i=1}^n (X_i - \overline{X})^2 \right]$.

We add and subtract the true population mean $\mu$ inside the summation:
$$\sum_{i=1}^n (X_i - \overline{X})^2 = \sum_{i=1}^n ((X_i - \mu) - (\overline{X} - \mu))^2$$

Expanding the squared term $(a-b)^2 = a^2 - 2ab + b^2$:
$$= \sum_{i=1}^n \left( (X_i - \mu)^2 - 2(X_i - \mu)(\overline{X} - \mu) + (\overline{X} - \mu)^2 \right)$$

Distributing the summation:
$$= \sum_{i=1}^n (X_i - \mu)^2 - 2(\overline{X} - \mu) \sum_{i=1}^n (X_i - \mu) + \sum_{i=1}^n (\overline{X} - \mu)^2$$

Note that $\sum_{i=1}^n (X_i - \mu) = n(\overline{X} - \mu)$ and
$\sum_{i=1}^n (\overline{X} - \mu)^2 = n(\overline{X} - \mu)^2$.
Substituting these back:
$$= \sum_{i=1}^n (X_i - \mu)^2 - 2n(\overline{X} - \mu)^2 + n(\overline{X} - \mu)^2$$
$$= \sum_{i=1}^n (X_i - \mu)^2 - n(\overline{X} - \mu)^2$$

**Applying Expectation** Now we take the expected value of the
simplified expression:
$$E\left[ \sum_{i=1}^n (X_i - \overline{X})^2 \right] = \sum_{i=1}^n E[(X_i - \mu)^2] - n E[(\overline{X} - \mu)^2]$$

Recall the definitions of population variance and variance of the sample
mean: 1. $E[(X_i - \mu)^2] = Var(X) = \sigma^2$ 2.
$E[(\overline{X} - \mu)^2] = Var(\overline{X}) = \frac{\sigma^2}{n}$

Substituting these values:
$$= \sum_{i=1}^n \sigma^2 - n \left( \frac{\sigma^2}{n} \right)$$
$$= n\sigma^2 - \sigma^2$$ $$= (n-1)\sigma^2$$

**Expected Values of the Estimators**

1.  For the **Biased Estimator** ($\sigma_n^2$):
    $$E[\sigma_n^2] = E\left[ \frac{1}{n} \sum_{i=1}^n (X_i - \overline{X})^2 \right] = \frac{1}{n} (n-1)\sigma^2 = \frac{n-1}{n}\sigma^2$$
    Since $E[\sigma_n^2] \neq \sigma^2$, this estimator is biased.

2.  For the **Unbiased Estimator** ($\sigma_{n-1}^2$):
    $$E[\sigma_{n-1}^2] = E\left[ \frac{1}{n-1} \sum_{i=1}^n (X_i - \overline{X})^2 \right] = \frac{1}{n-1} (n-1)\sigma^2 = \sigma^2$$
    Since $E[\sigma_{n-1}^2] = \sigma^2$, this estimator is unbiased.

## Task f)

An estimator $\hat{\theta}$ is defined as unbiased if its expected value
equals the true parameter $\theta$:

$$E[\hat{\theta}] = \theta \quad \iff \quad Bias(\hat{\theta}) = E[\hat{\theta}] - \theta = 0$$ 1.
Testing the Estimator $\sigma_n^2$

From part (e), we have:

$$E[\sigma_n^2] = \frac{n-1}{n}\sigma^2$$ Calculating the Bias:

$$Bias(\sigma_n^2) = E[\sigma_n^2] - \sigma^2 = \frac{n-1}{n}\sigma^2 - \sigma^2 = \sigma^2 \left( \frac{n-1}{n} - 1 \right) = \sigma^2 \left( \frac{n-1 - n}{n} \right) = \sigma^2 \left( \frac{-1}{n} \right) = -\frac{\sigma^2}{n} $$

Since $Bias(\sigma_n^2) \neq 0$, the estimator $\sigma_n^2$ is biased
(it systematically underestimates the variance).

2.  Testing the Estimator $\sigma_{n-1}^2$

From part (e), we have:

$$E[\sigma_{n-1}^2] = \sigma^2$$

Calculating the Bias:

$$Bias(\sigma_{n-1}^2) = E[\sigma_{n-1}^2] - \sigma^2 = \sigma^2 - \sigma^2 = 0$$
Since $Bias(\sigma_{n-1}^2) = 0$, the estimator $\sigma_{n-1}^2$ is
unbiased.

## Task g)

The results from our task c) are in perfect agreement with the
theoretical derivations (Tasks e and f).

Theory: we proved mathematically that $E[\sigma_{n-1}^2] = \sigma^2$ and
$E[\sigma_n^2] = \frac{n-1}{n}\sigma^2$.

Practice: The simulation showed that the average of $\sigma_{n-1}^2$
over 10,000 runs was almost exactly 4 (the true variance), while the
average of $\sigma_n^2$ was consistently lower (e.g., $\approx 3.6$ for
$n=10$), matching the theoretical prediction of systematic
underestimation.

The fundamental reason $\sigma_n^2$ is biased is that it treats the
sample mean $\overline{X}$ as if it was the true population mean $\mu$.
However, $\overline{X}$ is calculated from the data itself, which
minimizes the sum of squared deviations for that specific sample. This
makes the deviations $(X_i - \overline{X})^2$ slightly smaller on
average than $(X_i - \mu)^2$. Dividing by $n-1$ (degrees of freedom)
instead of $n$ perfectly compensates for this.

Practical Implication: Small Samples: For small datasets (e.g.,
$n < 30$), using the biased estimator $\sigma_n^2$ leads to a
significant error. In these cases, using $\sigma_{n-1}^2$ is critical.

Large Samples: As $n \to \infty$, the difference between $\frac{1}{n}$
and $\frac{1}{n-1}$ becomes negligible. Both estimators converge to the
true parameter, making $\sigma_n^2$ asymptotically unbiased.
